Viscous Erosion in Porous Media

### 2018-01-15 docs/methods/erosion_methods.tex
NM

For the sake of numerical stability, we modify the erosion law in
equation~\eqref{eqn:bodyEvolution} to include a term that depends on the
local curvature $\kappa$. The idea is to prevent regions of very high
curvature. The modified normal velocity is 
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{align}
Here, $\eps$ is a small parameter, $L$ is the total arc length of the
body, and $\mean{\cdot}$ indicates a boundary mean. The second term on
the right acts to smooth bodies with high-frequency oscillations by
introducing a {\em curvature}-dependent component in the interface
velocity. Since a closed body has a mean curvature of $\mean{\kappa} =
2\pi / L$, the term in parentheses has mean zero. Thus, this term is
area-preserving, so that the only material loss is due to the shear
stress itself. In addition, we have scaled this term on the
$\mean{\atau}$, the average absolute stress, so that it is always
proportional to the true source of erosion $\atau$, even as bodies
vanish \todo[inline]{reword}.

@@

For simplicity, we introduce the variables
\begin{align*}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atausig}}{L}.
\end{align*}
Then, the evolution equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
Here $\NLterm$ represents the nonlinear terms (and nonlocal terms) in
the PDE for $\theta$.

We can evolve $\theta$ in spectral space since $\theta(\alpha)$ is
periodic up to a jump of $2\pi$ for $\alpha \in [0,2\pi]$.  Therefore,
we introduce the Fourier series
\begin{align}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha}, \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha}.
\end{align}
The careful reader will notice that we have separated out the linear
part of $\theta$, so that the remaining function is smooth and has a
rapidly decaying Fourier series.  In spectral
space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k.
\end{align*}
Each ODE is first-order and linear (though variable-coefficient due to
$\elfun$) and thus can be solved by using the integrating factor
\begin{align}
  \label{eq:mu}
  \mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right).
\end{align}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k.
\end{align}
\todo[inline]{multiply defined references here}
We solve system~\eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we
choose the {\em midpoint} RK2 method so as to maximize the smoothing
effect of the curvature term.

At the first stage of RK2 (the half step), we discretize
equations~\eqref{eqn:Levo2}--\eqref{eqn:thetaevo2} as
\begin{align}
\label{eq:L12}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = \Mterm^{(n)} \\
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \thhat_k^{(n)}}{\Dt/2} = \hat{\NLterm}_k^{(n)}
\end{align}
where the superscript indicates the time step. For simplicity, we have
chosen the constant of integration in the exponent of
equation~\eqref{eq:mu} so that $\mu_k^{(n)} = 1$. At the next stage of
RK2, we must compute $\Mterm^{(n+1/2)}$ and
$\widehat{\NLterm}_k^{(n+1/2)}$, which requires a call to the Stokes
solver. Hence, we must compute $\mu_k^{(n+1/2)}$, so that
$\thhat_k^{(n+1/2)}$ can be recovered and the geometry at the mid-step
can be computed. To compute $\mu_k^{(n+1/2)}$, we use the left-end-point
rule (first order, consistent with first forward Euler step of RK2)
\begin{align*}
\mu_k^{(n+1/2)} = \exp \left( \epsilon k^2 \frac{\Dt}{2}
\elfun^{(n)}\right).
\end{align*}
It is instructive to recast as
\begin{align*}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{\NLterm}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
\end{align*}
so that the smoothing effect of the exponential is obvious. With
$\thhat_k^{(n+1/2)}$ computed, it is now possible to compute
$\Mterm^{(n+1/2)}$ and $\widehat{\NLterm}_k^{(n+1/2)}$.

@@

The layer potential~\eqn{eqn:genericLP} is approximated with the
trapezoid rule 
\begin{align}
  \phi(\xx) = \sum_{j=1}^N q_j K(\xx,\yy_j),
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density
function.  Based on numerical experiments, the error of the $N$-point
trapezoid rule is acceptable for points $\xx$ that are more than 5
arclength spacings from $\bd\Omega$.

@@

With the description of the methods complete, we now present numerical results on bodies eroding in Stokes flow. We first discuss the case of a single eroding body, as certain features can be predicted analytically and thus used for validation. We then discuss the erosion of multiple bodies and, in particular, important differences that arise as compared to the single-body case.

@@

Figure \ref{01bodseq} shows the erosion of a single body, numerically simulated with $N_\iin = 1024$ discretization points. The figure shows the body at three evenly spaced times during the erosion process, with the color map to indicate the vorticity of the surrounding flow. When evaluated on the boundary, the vorticity indicates the shear stress and thus the local erosion rate. As seen in the figure, the body is initially circular, with strongest vorticity above and below it. The flow is for-aft symmetric as a result of the time-reversibility of the Stokes equations, and vorticity vanishes at the body's front and rear stagnation points. At the next time step, the body has lost material due to erosion, but remains for-aft symmetric due to the flow symmetry. The greatest material loss has occurred at the body's top and bottom, where shear is strongest. Interestingly, the body has developed well-defined corners at its front and rear stagnation points. This geometric feature is a consequence of the erosion law, which depends on the {\em absolute} shear stress: wherever the stress vanishes, $\atau$ exhibits a corner, which manifests as a corner in the geometry. As shown in the third snapshot, this corner persists as the bodies becomes even more slender while vanishing.

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Convergence test: the stopping time is $t_s =$ 1E-2, which is 56\% of the vanishing time.
GOOD TERM: Cauchy-convergence.} 
\caption{Convergence test: the stopping time is $t_s =$ 1E-2, which is
56\% of the vanishing time.
} 
\vspace{0.3 pc}
\label{convtab}
\begin{tabular}{c l l}
\hline
\hspace{0.5pc} $\Delta t/t_s$
\hspace{0.5pc} & error 
\hspace{0.5pc} & order \\
\hline
1/100	& 4.69E-6		& --		\\
1/200	& 1.26E-6		& 1.90	\\
1/400	& 3.26E-7		& 1.95	\\
1/800	& 8.28E-8		& 1.98	\\
1/1600	& 2.09E-8		& 1.99	\\
1/3200	& 5.26E-9		& 1.99	\\
1/6400	& --			& --		\\
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%

>>>>

For the sake of numerical stability, we modify the erosion law in
equation~\eqref{eqn:bodyEvolution} to include a term that depends on the
local curvature $\kappa$. The idea is to prevent regions of very high
curvature. The modified normal velocity is 
\begin{align}
  \label{eqn:Vn}
  \Vn = \atau + \eps \mean{\atau} \left( \frac{L }{2\pi} \kappa - 1 \right)
\end{align}
Here, $\eps$ is a small parameter, $L$ is the total arc length of the
body, and $\mean{\cdot}$ indicates a boundary mean. The second term on
the right acts to smooth bodies with high-frequency oscillations by
introducing a {\em curvature}-dependent component in the interface
velocity. Since a closed body has a mean curvature of $\mean{\kappa} =
2\pi / L$, the term in parentheses has mean zero. Thus, this term is
area-preserving, so that the only material loss is due to the shear
stress itself. In addition, we have scaled this term on the
$\mean{\atau}$, the average absolute stress, so that it is always
proportional to the true source of erosion $\atau$, even as bodies
vanish. \todo[inline]{reword}

@@

For convenience, we introduce the variables
\begin{align*}
%\label{eq:M}
& \Mterm = - 2 \pi \mean{\thalpha \Vnsig}, \\
%\label{eq:N}
& \NLterm = \frac{2 \pi}{L} \left( \pderiv{\atausig}{\alpha} + \thalpha
\Vssig \right), \\
%\label{eq:elfun}
& \elfun = \frac{2 \pi  \mean{\atausig}}{L}.
\end{align*}
Then, the evolution equations~\eqref{eqn:Levo1}
and~\eqref{eqn:thetaevo1} can be compactly expressed as
\begin{align}
\label{eqn:Levo2}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo2}
& \pderiv{\theta}{t} = \eps \elfun \ppd{\theta}{\alpha} + \NLterm.
\end{align}
Here $\NLterm$ represents the nonlinear terms (and nonlocal terms) in
the PDE for $\theta$.

We can evolve $\theta$ in spectral space since $\theta(\alpha)$ is
periodic up to a jump of $2\pi$ for $\alpha \in [0,2\pi]$.  Therefore,
we introduce the Fourier series
\begin{align}
 \theta(\alpha,t) &= \alpha + \FourierSum \thhat_k e^{ i k \alpha}, \\
 \NLterm(\alpha,t)  &= \FourierSum \widehat{\NLterm}_k e^{ i k \alpha}.
\end{align}
The careful reader will notice that we have separated out the linear
part of $\theta$, so that the remaining function is smooth and has a
rapidly decaying Fourier series.  In spectral
space,~\eqref{eqn:thetaevo2} produces the system of ODEs
\begin{align*}
\tderiv{\thhat_k}{t} +  \eps k^2  \elfun \thhat_k = \widehat{\NLterm}_k.
\end{align*}
Each ODE is first-order and linear (though variable-coefficient due to
$\elfun$) and thus can be solved by using the integrating factor
\begin{align}
  \label{eq:mu}
  \mu_k = \exp \left( \eps k^2 \int \elfun \, dt \right).
\end{align}
We are therefore led to the coupled system of ODEs
\begin{align}
\label{eqn:Levo3}
& \tderiv{L}{t} = \Mterm, \\
\label{eqn:thetaevo3}
& \tderiv{}{t}\left( \mu_k \thhat_k \right) = \mu_k \widehat{\NLterm}_k.
\end{align}
We solve system~\eqref{eqn:Levo3}--\eqref{eqn:thetaevo3} with an
explicit, second-order Runge-Kutta method (RK2). In particular, we
choose the {\em midpoint} RK2 method so as to maximize the smoothing
effect of the curvature term.

At the first stage of RK2 (the half step), we discretize
equations~\eqref{eqn:Levo3}--\eqref{eqn:thetaevo3} as
\begin{align}
\label{eq:L12}
& \frac{L^{(n+1/2)} - L^{(n)}}{\Dt/2} = \Mterm^{(n)} \\
& \frac{ \mu_k^{(n+1/2)} \thhat_k^{(n+1/2)} - \thhat_k^{(n)}}{\Dt/2} = \hat{\NLterm}_k^{(n)}
\end{align}
where the superscript indicates the time step. For simplicity, we have
chosen the constant of integration in the exponent of
equation~\eqref{eq:mu} so that $\mu_k^{(n)} = 1$. At the next stage of
RK2, we must compute $\Mterm^{(n+1/2)}$ and
$\widehat{\NLterm}_k^{(n+1/2)}$, which requires a call to the Stokes
solver. Hence, we must compute $\mu_k^{(n+1/2)}$, so that
$\thhat_k^{(n+1/2)}$ can be recovered and the geometry at the mid-step
can be computed. To compute $\mu_k^{(n+1/2)}$, we use the left-end-point
rule (first order, consistent with first forward Euler step of RK2)
\begin{align*}
\mu_k^{(n+1/2)} = \exp \left( \epsilon k^2 \frac{\Dt}{2}
\elfun^{(n)}\right).
\end{align*}
It is instructive to recast as
\begin{align*}
\thhat_k^{(n+1/2)} = \left( \thhat_k^{(n)} + \frac{1}{2} \Dt \, \hat{\NLterm}_k^{(n)} \right)
\exp \left( -\epsilon k^2 \frac{\Dt}{2} \elfun^{(n)}\right)
\end{align*}
so that the smoothing effect of the exponential is obvious. With
$\thhat_k^{(n+1/2)}$ computed, it is now possible to compute
$\Mterm^{(n+1/2)}$ and $\widehat{\NLterm}_k^{(n+1/2)}$.

@@

The layer potential~\eqref{eqn:genericLP} is approximated with the
trapezoid rule 
\begin{align}
  \phi(\xx) = \sum_{j=1}^N q_j K(\xx,\yy_j),
  \label{eqn:layerPotential}
\end{align}
where $q$ contains the quadrature weights, Jacobian, and density
function.  Based on numerical experiments, the error of the $N$-point
trapezoid rule is acceptable for points $\xx$ that are more than 5
arclength spacings from $\bd\Omega$.

@@

With the description of the methods complete, we now present numerical results on bodies eroding in Stokes flow. We first discuss the case of a single eroding body, as certain features can be predicted analytically and thus used for validation. We then discuss the erosion of multiple bodies and highlight important differences that arise as compared to the single-body case.

@@

Figure \ref{01bodseq} shows the erosion of a single body, numerically simulated with $N_\iin = 1024$ discretization points. The figure shows the body at three evenly spaced times during the erosion process, with the color map to indicate the vorticity of the surrounding flow. When evaluated on the boundary, the vorticity is equivalent to shear stress and thus indicates the local erosion rate. As seen in the figure, the body is initially circular, with strongest vorticity above and below it. The flow is for-aft symmetric as a result of the time-reversibility of the Stokes equations, and vorticity vanishes at the body's front and rear stagnation points. At the next time step, the body has lost material due to erosion, but remains for-aft symmetric due to the flow symmetry. The greatest material loss has occurred at the body's top and bottom, where shear is strongest. Interestingly, the body has developed well-defined corners at its front and rear stagnation points. This geometric feature is a consequence of the erosion law, which depends on the {\em absolute} shear stress: wherever the stress vanishes, $\atau$ exhibits a corner, which manifests as a corner in the geometry. As shown in the third snapshot, this corner persists as the bodies becomes even more slender while vanishing.

@@

% Table
%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{table}%[htbp]
\begin{center}
\caption{Convergence test: the stopping time is $t_s =$ 1E-2, which is 56\% of the vanishing time.
GOOD TERM: Cauchy-convergence.} 
\caption{Convergence test: the stopping time is $t_s =$ 1E-2, which is
56\% of the vanishing time.
} 
\vspace{0.3 pc}
\label{convtab}
\begin{tabular}{c l l}
\hline
\hspace{0.5pc} $\Delta t/t_s$
\hspace{0.5pc} & error 
\hspace{0.5pc} & order \\
\hline
1/100	& 4.69E-6		& --		\\
1/200	& 1.26E-6		& 1.90	\\
1/400	& 3.26E-7		& 1.95	\\
1/800	& 8.28E-8		& 1.98	\\
1/1600	& 2.09E-8		& 1.99	\\
1/3200	& 5.26E-9		& 1.99	\\
1/6400	& --			& --		\\
\hline
\end{tabular}
\end{center}
\end{table}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Parameters in this test are: 01circ256, nits = 7, dt0 = 1e-4, tfin = 1e-2
% epsfac = 10, sigfac = 10, fixarea = 0, fixprdop = 0
